A Guide to Designing Experiments to Test Statistical Graphics

### 2025-03-17 response.qmd

---
title: "Response to Review"
format: pdf
bibliography: refs.bib
---

We thank the reviewer for their careful notes about our manuscript. 
We have addressed many of the issues identified; in some cases, we defer to Dr. Jurgen Symanczyk, who asked that we e.g. not focus as much on statistical lineups because there is at least one other paper slated for WIRE on that topic which will provide much more detail.

Below, please find the point-by-point responses.

- The title refers to “statistical graphics”, however in text authors use “charts”. In fact
on Page 1, line 27: Authors write “charts and data visualization” but why the
distinction?

> In our experience, graphics, charts, and visualizations are used nearly interchangeably, and we often rotate between them to ensure that no single word becomes semantically satiated.

- The authors call the visualization lineup protocol as “statistical lineups” but is this really a good name for it? It has not been called as such before and I don’t think it is a good idea to give unintuitive names (it will confuse future research in the area). If you must give it a name, I would prefer “visualization lineups” but it should really be defined, e.g. “In this paper, visualization lineups refer to ...".

> The term statistical lineups have been used throughout several papers, including @loyDiagnosticToolsHierarchical2013,@vanderplasSpatialReasoningData2016,@loyModelChoiceDiagnostics2017, @vanderplasTestingStatisticalCharts2020, @vanderplasStatisticalSignificanceCalculations2021, @liPlotWorthThousand2024, @robinsonPerceptionCognitiveImplications. 
> We have added some text explaining the origin of the term.

- The tone is sometimes casual, e.g. Page 8, Line 33: “system is very, very good”, Page 13, Line 5: “a combination of density and rug geoms”. These sound colloquial and it is not of an academic convention to write things like “very, very” and “rug geoms”.

> We have addressed these language issues, but note that 'rug geom' is a convention from the grammar of graphics as implemented in ggplot2, describing the type of chart. We've adjusted how we refer to this to clarify that it is a composition of two different plots which show both the aggregate density and the individual observations.

- Citation style issues. E.g., page 8, Line 3 “Lu et al (Lu e al, 2022)” should be just “Lu
et al (2022)”.

> We've fixed these (whoops) and also the citation issue noted further below ([@shah]).

- Being specific, e.g. Page 1 line 43: be explicit that “test charts _with human
evaluation_”.

> We've used slightly different language (test the visualization on humans), but have added this specificity.

- I would appreciate a more informative caption for Figure 1.

>  The caption now reads: "Direct adjustment of a plot in a perceptual task. In this experiment designed to assess the strength of the sine illusion, the user adjusts the plot using - and + buttons, which control the strength of a transformation designed to correct the effect of the sine illusion. When the user is satisfied that the lines are of equal length, they select the ‘Finished’ button to move to the next task. The experiment used a psychophysics experimental design, the method of adjustment, but leveraged the interactive Shiny interface to record the entire sequence of adjustments made by the user for each trial. A demo version of this application can be found at https://shiny.srvanderplas.com/sine-illusion/."

- The description of the lineup itself and its role in visual testing is somewhat abstract. Figure 5 might need to be introduced earlier for better clarity. Similarly, no explanation to how the see-value is calculated – if this is expected knowledge, could authors give recommendation of what literature readers should read to know the appropriate background, otherwise I think it is useful to make this guide self-contained and describe it briefly.

> We specifically were asked to not go into depth about lineups because of a pair of other articles which are intended to be published alongside this article. We have added the appropriate references so that readers can find the calculation details there. We have also included some additional language (which can be removed easily if not desireable) referring readers to the companion article about lineups.

- Page 1 line 46: the concept of internal and external validity is not explained. This might be common knowledge in psychometrics or psychology but I don’t think it is in statistics.

> Well, that is a pity, as the concepts are broadly applicable to many different areas of statistics and were part of my statistical education. In any case, the language has been rephrased to refer to generalizability (external validity) and experimental control (internal validity), which hopefully will capture not only statisticians but also the InfoVis crowd made up of e.g. computer scientists and psychologists.

- Is there a guideline or reference for “think-aloud” processes? It’s a concept that not a lot of statisticians would be familiar with so it would be nice to know more about this and the benefits of this approach.

> We've added several additional citations for think-aloud processes and explained both the informal use in experiment testing and the formal use in data collection.

- Should you assume your audience knows what a “PHP” (page 11, line 18) is? Same goes for Qualtric surveys, Amazon Mechanical Turk, Prolific and Reddit. These have no reference/link and little to no explanation. Given technology platforms can evolve rapidly, perhaps authors would like to give more context so that their paper can be understood by future readers. E.g. hypothetically suppose if Reddit disappears today then researchers in 20 years may not know what Reddit is.

> I have attempted to explain these sites in context wherever possible and link to relevant information. I have emphasized that Reddit is a social media site alongside X, Mastodon, and Bluesky -- while we have not yet attempted to pilot studies on Mastodon/Bluesky, those are the natural replacements for X now that it is becoming less popular. I have screenshots of a recruiting message on X as well as the reddit.com/r/samplesize homepage, which could be included in the paper but seem slightly distracting. I have linked to an Internet Archive of the samplesize homepage to preserve this information for the future, but X blocks archival sites, so a similar archive could not be created for that platform. If the goal is that the paper is still comprehensible in 20 years, I think this should be sufficient - MySpace and Blogger, which have been obsolete for 15+ years, still have wikipedia pages which could be used to understand the references.

- What is IRB? Page 10, Line 43: “make the argument to IRB that our research is exempt”.

> We've added the acronym definition and some additional information clarifying the ethics review process for human subjects experiments, in response to an additional suggestion.

- Also NORC panel is not defined.

> This one is surprisingly tricky - NORC used to stand for National Opinion Research Center, but they got rid of the acronym and just named the organization NORC within the last few years. We've clarified that it's an organization like Gallup and that the basic goal is to have a panel survey rather than a sample recruited from a platform like Prolific/MTurk.

- Rasch models on Page 12, Line 39 not defined nor cited.

> We've added additional citations for Rasch models, but as they're suboptimal relative to mixed models for these types of experiments, we hope that this will be sufficient, as we don't want to spend the space explaining a less powerful model.

- how many participants should be recruited?

> This is so heavily dependent on the experimental design and factors under investigation that it is difficult to offer guidance on sample size. In attempting to add information about sample size, we refactored the paper somewhat, differentiating the protocol development process from the experimental design calculations. Graphics studies are sometimes overwhelming because statisticians do not usually control both the subject matter expert decisions and the statistical design decisions, and separating these two out helps to ensure that power calculations and statistical design decisions are isolated (as much as possible) from protocol decisions like the number of trials per participant, which is typically determined based on fatigue.

- what number of plots should be shown to participants in a lineup?

> We have attempted to not make this paper specific to lineup studies, so we will defer this discussion to the visual inference paper that is also to be submitted to WIRE.

- mention of ethic approval? A reminder that people do need to have ethics approval lined up and linked in their study to use results for publication?

> A section on ethics approval has been added, though we have no experience with this process outside of the United States and so have tried to balance specificity (e.g. exempt research and expedited review) with the fact that many readers will be outside of the US and subject to different regulations.

- can authors comment on financial costs related to participant recruitment?

> Every platform is different and even within platforms, pricing schemes vary by the week or month. We've included some mention of considerations like participant completion time vs. number of participants with regard to platform fees, and have also included some general comments about the relative expense of panel surveys vs. online platforms vs. in-person studies vs. social media recruitment.

- any guide to ensure representative sample in participant recruitment? Do authors adopt particular method or strategies?

> We have added some additional references on the effects of participant sampling on results of studies, as well as the different characteristics of different platforms, though at least one of the studies we cite are published by those platforms, so there are conflict of interest concerns as well.

- what about filtering out obviously non-serious participants?

> A section on response validation and attention checks has been added

- Authors mention usage of Amazon Mechanical Turk, Prolific, Reddit and social media. Could they provide advice on which online recruitment platform to use?

> Additional discussions of the advantages of categories of recruitment platforms has been added, but we have tried to remain as factual as possible about different competitors within those categories, in part because this space is evolving rapidly and if mentioning sites like Reddit is a concern (as above), then recommendations for specific platforms would be similarly problematic. Some discussion of the evolution of platforms over time is included, though, to explain why Amazon Turk is less popular now than it was in the 2010s.

- any advice for the design of interface that participants use?

> This is heavily dependent on the choice of data collection software. Bespoke solutions like Shiny and PHP web applications can be infinitely customized, but platforms like Qualtrics and Google Forms are much more rigid in design. Any advice we could give about the interface would have to handle many different configurations.

- Any examples of qualitative questions asked?

> We have added a few suggestions for qualitative questions, and reorganized the qualitative analysis section to better emphasize the citation to @clusters, which contains a qualitative analysis that complements the quantitative modeling.

- Page 1 line 36: grammar of graphics technically doesn’t “classify” charts but just is a means to specify one.

> Yes, this is true, but the grammar also provides a convenient way to classify charts that separates the mappings from the geometric representation. This is particularly powerful in graphics studies, where different geometrical representations of the same mappings can be compared head-to-head. We've modified the language slightly in the hopes of clarifying what we mean here.

- Page 2 line 46: “target plot with signal” — but there can be a target plot with no signal right?

> We've amended the language to avoid this issue.

-  “automatically transcribed using text-to-speech functions of large models” — why LLM particularly? What about the approaches that pre-dates LLM?

> It is only with the advent of LLMs that these features have reached an accuracy where speech-to-text might be a viable option for think-aloud recording, but we've removed the mention of LLMs in any case.

- Page 6, line 40: what is "(shah?)” supposed to be?

> That's a failed markdown reference that we have fixed. Thanks for catching it!

### 2025-03-17 index.qmd
Updating refs and finishing response

@@

# Experiment Development {#sec-exp-dev}

It would be difficult to develop a full data generating model without some idea of the experimental protocol: the basic equipment required for the experiment, some idea of what questions users will answer, where and how data will be collecteed, and so on. 
These experimental design factors are fairly natural for scientists to accumulate over the course of imagining and planning an experiment. 
When conducting graphical tests, however, there are additional considerations beyond those taught in a standard experimental design course.
Experimenters must carefully consider how much information participants should have about the experiment, what platform to use to recruit participants, and the experimental infrastructure underlying data collection. 
In addition to our brief overview of considerations in our research, @kochariConductingWebBasedExperiments2019 has many helpful suggestions for conducting web-based cognitive and perceptual experiments that also apply to studies of statistical graphics.

@@

### Infrastructure

@@

### Participant Recruitment

@@

### Participant Instructions

@@

### Attention Checks and Answer Validation

@@

- Attention checks - questions inserted into the experiment which appear similarly to actual trials but which instruct participants to select a specific response or complete a trivial task (e.g. select the result of 2+2). More guidance on attention checks can be found in @muszynskiAttentionChecksHow2023.

- User input validation - mechanisms which reduce the potential parameter space of user inputs to a valid or reasonable set of parameters. Examples include not allowing participants to select negative variance values, requiring continuous user-drawn lines on an interactive chart, and even checking to ensure that the user's input is of the correct type.

-

### Demographic Data

@@

# Experimental Design Considerations {#sec-exp-design}

In this section, we broadly consider three different components of the experimental design process: infrastructure for data collection and participant recruitment, experimental design, and ethics panel review.

## Data Collection

## Experimental Design

Data collection process and infrastructure considerations can be complex in graphics experiments, but they are critically important for ensuring that data is successfully collected and that results can be generalized to the wider population. 
However, there are also several considerations that are related to more standard experimental design concerns, but with twists unique to cognitive science experiments. 
Experimenters must consider factors like cognitive load when determining how to block and randomize factors and allocate trials to participants. 
These factors are also extremely important when deciding on the number of participants necessary for an experiment.

## Cognitive Load

In our experience, it is difficult to expect participants to evaluate more than about 15 charts in one sitting. 
If users are asked to deeply engage and answer multiple questions about each chart, this limit may be lower (8-10), but even with a relatively simple task, as in lineup methods, it is hard for participants to accurately evaluate more than about 15 charts. 
Tasks which are more interactive, such as `You Draw It', may be somewhat easier for participants, but it is unlikely that participants would be willing to complete more than about 20 tasks in one sitting even with tasks that require fewer decisions because these tasks often require more engagement. 
At some point, participants' interest in accuracy or attention to detail will decline, and the researcher is better off using a larger number of participants completing fewer tasks within the window where participants are attentive and engaged.
As a result, constraints on cognitive load are perhaps the most important factor in determining the sample size necessary for a study.

## Sample Size

It is difficult to offer specific advice on the number of participants to include in the study, because the experimental design, participant engagement level, and the specific factors under investigation have such a large effect on the number of trials one participant can reasonably complete before fatigue effects increase response variance. 
Because there are so many factors which have an effect on the required sample size, explicit power calculations for these experiments can be extremely difficult - experimenters would have to run each experiment once to get the necessary information for a power calculation and then run the **exact same** experiment a second time with the calculated $n$ participants.
Often, after the pilot experiment is completed, more significant changes are necessary that invalidate the estimates which might contribute to the power calculation.
Some experts have begun to recommend that instead of pilot-study informed power calculations, experimenters use a more general approach based on the intended analysis method [@brysbaertHowManyParticipants2019].

When designing graphics experiments, we typically optimize sample size according to constraints identified during the iterative development of the experiment's structure.
First, we optimize the model and data generation method.
Next, we revise the graphics used in the experiment, removing any extraneous information and ensuring that sufficient cues (ticks, grid lines, labels, colors, aesthetics) are provided across experimental conditions to complete the required graphical task.  
Then, we develop the data collection application and determine what types of questions the user will answer during the experiment.
This includes both the primary questions of interest and any secondary measures, such as completion time, participant reasoning, confidence level, and so on. 
As the number of tasks required of the participant increases, the number of trials we can ask participants to complete during a session decreases. 
We pilot test the application, often using a pool of graduate students, colleagues, social media acquaintances, and conscripted family members. 
During the pilot test stage, we pick an optimistic number of trials per participant -- for instance, we might ask pilot participants to evaluate 20 graphical lineups (400 total plots), even though we expect to reduce the number to 10 or 15 during the main experiment. 
After pilot participants complete the study, we ask questions about fatigue to determine what range of trials per participant is reasonable; in order to find this point, pilot testing needs to include more trials than we expect people to complete without fatigue effects.

At the end of our pilot study, the power calculation becomes an optimization of constraints imposed by the experimental design (e.g. how many blocks in an incomplete block design can we have with the factors under investigation), anticipated analysis method, and budget for the experiment. 
The number of trials a single participant can complete without fatigue affects the block size and strategy by which we allocate trials to each participant.
In addition, we must consider how long participants take to complete the required number of trials.
Completion time is used to determine participant compensation (if using a participant recruiting platform). 
Ethics boards and some recruitment platforms require that participants are paid a reasonable wage for their time (currently, around \$15 US per hour), and platforms may ask for median completion time and automatically reject submissions from participants who are too far under or over the specified time limits; they may also require additional participant payments if the median time estimate is too far below the actual average completion time during the experiment.
Platforms may also calculate fees based on both the participant payment and number of participants recruited, with additional fees to recruit e.g. demographically representative samples; as a result, it can be advantageous to balance cognitive load concerns with the fee structure used by the selected recruitment platform.
This complex calculus is often not considered in sample size calculations taught in statistics courses, but experimenters do not often have budgets of infinite size to complete experiments, so some financial optimization is necessary in practice. 
This pragmatic sample size calculation may boil down to designing the experiment in a way that allows for collection of data using as many participants as the budget will support.

In psychophysical graphics experiments, such as @luModelingJustNoticeable2022, there are often fewer participants with many more trials per participant (28 participants, 250 trials each); the stimuli in these experiments are often much simpler (e.g. one or two plots instead of 20 in a lineup) and engagement is limited to detection rather than estimation or prediction.
Psychophysics experiments, particularly those designed for analysis using Rasch models, require that each participant assesses a full factorial set of stimuli, whereas analysis with generalized linear mixed models allows for use of incomplete blocks and other strategies that spread the cognitive burden across several participants. 
Online experiments which require more complex engagement with statistical graphics often have between 300 and 600 participants [@robinsonPerceptionCognitiveImplications,], but this varies widely with the experimental design, type of stimuli used, and level of participant engagement required; @loyVariationsQQPlots2016 and @clusters had more than 1300 participants,  while @sineillusion and @vanderplasFramedReproducingRevisiting2019 had fewer than 150 participants. 
In-person experiments [e.g. @vanderplasSpatialReasoningData2016] often use fewer participants than online equivalents, but these experiments may require more tasks of each participant, which reflects the increased logistical costs of scheduling participants, supervising the experiment, and entering the data, if tasks are completed on paper. 
In some cases, in-person experiments may combat cognitive fatigue by scheduling multiple visits per participant; this leaves experiments vulnerable to participant drop-out effects, but can effectively balance cognitive load and participant recruitment costs in some situations.

## Trial Allocation to Participants

In statistical design terms, most of our studies involve some type of balanced incomplete block design, where participants are assigned to a subset of experimental conditions which allow for estimation of the full range of effects specified in the model. 
The particular structure of these designs depends heavily on the factorial structure of the study, but we typically arrange participants' trials to ensure that participants see the same data set only once (where possible) and see as many different experimental conditions as possible. 
It is also important to reduce the impact of order effects using either e.g. Latin square designs or randomization where possible, but we recognize that this is not always feasible due to the need to maintain participant naivete during some portions of the experiment.

## Ethical Review of Human Subjects Experiments

In almost any research environment, there are requirements that experiments involving human participants undergo review by an ethics board to ensure that participants are not harmed by participating in the study.
We are most familiar with regulations in the United States and can only speak to that environment.
Most graphics experiments conducted in the US fall under the "exempt" category of experiments which require only basic review and approval. 
These experiments record only basic demographic information, ask participants to complete tasks that involve no risk, and experimenters record information which would not be embarrassing to participants if exposed. 
In order to ensure that our experiments fall into this category, we will often ensure that demographic information is collected and stored separately from any identifiable information, such as e.g. user IDs from the participant recruitment platform that allow us to monitor task completion and pay participants for their time.
Researchers should also ensure that they comply with privacy laws such as the European [General Data Protection Regulation (GDPR)](https://gdpr-info.eu/chapter-3/); complying with this law while collecting data which is not identifiable can be complicated. 
We recommend consulting with your ethics board and institutional recommendations in order to maintain legal compliance and safeguard participant privacy.

@@

Once the data generating model is set, the experiment is designed, and the charts have been developed, the next step is an extensive round of pilot testing. 
The goal of pilot testing is to ensure that the experiment is set up properly and that no issues have been overlooked. Pilot testing also provides an opportunity to ensure that directions are clear, participants know what they are supposed to be doing, and to estimate task completion time (which determines how much participant will be paid on many online testing platforms). 
Our studies usually go through 2-3 rounds of pilot testing, with at least one of those rounds involving any relatives and friends who are less technologically savvy. 
We also purposely include talented individuals who can accidentally crash any testing applications. 
A final round of testing usually includes any and all coworkers and friends who may be available for 10-15 minutes during the work day. 
Pilot study samples do not need to be representative of any particular population, though we would caution against using exclusively visualization researchers or statisticians in your pilot sample because some issues are less likely to show up with knowledgeable participants. 
In some studies, each participant may see a new set of data, depending on how the study is designed. 
In these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine exactly what happened and/or how things went wrong. 
Hard drive space is extremely cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.

One highly useful (but not strictly essential) component of an experiment that can be set up during the pilot testing stage is a basic analysis script which summarizes all data collected to date visually. 
We have used such scripts in the past to produce automatically updating dashboards or web pages, allowing for real time or near-real time monitoring of data collection efforts. 
This provides an easy way to summarize completion of the experiment so that individuals can receive credit (if using services like MTurk or Prolific), but also allows interested participants to see the data if recruiting participants who are more likely to be interested in the results, as sometimes happens on social media sites. 
As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can provide an illusion of control over the deluge of data, allowing any issues to be spotted and resolved relatively quickly. 
If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time. Prolific has recently added rate-limiting functions to the experimental control  platform, which makes controlling the number of active jobs much easier.].

@@

## Qualitative Responses
In many cases, it is helpful to combine participants' qualitative reasoning with their quantitative responses to designed graphical experiments. 
This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred. 
We have used word clouds to show overall themes in participant responses successfully in [@clusters]; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue. 
In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed (but analyzing this data quantitatively can be difficult).

>>>>

# Experiment Development Life Cycle

Developing a graphics experiment is often a highly iterative process, but it can help to approach the design process by first optimizing the model and data generation method before spending time on optimizing the specific stimuli or customizing the data collection platform.
This is important because the model parameters and data generation process inform the experiment structure and thus impact decisions made downstream.

Once the model and data generating mechanism are set, it is useful to revisit the primary questions of interest and determine how to measure the responses effectively. 
Secondary measures, such as response time, free responses, and confidence level should also be determined. 
These choices will inform the choice of a data collection platform and may also inform the participant recruitment method.

Next, we recommend developing a preliminary data analysis plan, specifying the general category of model which will be used (e.g. generalized linear mixed-effects model, t-test) and the contrasts which are most interesting. 
This sets up the experimental design decisions, but also ensures that as the data collection platform and process is developed, any design constraints are considered. 
Development of the data collection application is the next step, using draft graphics and the set of participant response measures of interest.

There are at least 3 stages of testing in a graphics experiment: informal tests, a pilot study, and the main experiment. 
The informal tests are critical for identifying issues with the data collection application, but can also be used to calibrate the number of tasks required of each participant. 
As the number and complexity of tasks increases, the number of trials we can ask participants to complete during a session decreases. 
The informal testing stage allows researchers to consider the tradeoffs inherent in the decision to reduce the amount of information collected for each task, reduce the number of tasks, or mitigate participant fatigue in other ways.

During preliminary testing, we use an optimistic number of trials per participant, so that we can determine when participants become overly fatigued. 
For instance, we might ask test participants to evaluate 20 graphical lineups (400 total plots), even though we expect to reduce the number to 10 or 15 during the main experiment. 
We test the application in individual or focus group sessions, often using graduate students, colleagues, social media acquaintances, and conscripted family members. 
After these participants complete the study, we ask questions about fatigue to determine what range of trials per participant is reasonable. 
At the end of our initial experiment tests, we have enough information to determine the basic parameters of the experimental design (e.g. how many blocks in an incomplete block design can we have with the factors under investigation). 
The number of trials a single participant can complete without excessive fatigue impacts the number of blocks and the strategy by which we allocate trials to each participant.

In addition, we must consider how long participants take to complete the required number of trials.
Completion time is used to determine participant compensation (if using a participant recruiting platform). 
Ethics boards and some recruitment platforms require that participants are paid a reasonable wage for their time (currently, around \$15 US per hour), and platforms may ask for median completion time and automatically reject submissions from participants who are too far under or over the specified time limits; they may also require additional participant payments if the median time estimate is too far below the actual average completion time during the experiment.
Platforms may also calculate fees based on both the participant payment and number of participants recruited, with additional fees to recruit e.g. demographically representative samples; as a result, it can be advantageous to balance cognitive load concerns with the fee structure used by the selected recruitment platform.

The findings from the initial test of the experimental procedure are then used to revise the data collection procedure in preparation for one or more pilot studies.
It is important to ensure that the software platform, trial allocation, and other components of the experiment are functioning as desired before a formal pilot study is conducted. 
In some cases, the pilot study is as simple as a "soft launch" of the main experiment, where the total number of trials is pre-specified and only a few trials are released initially to ensure that data collection works as expected. 
In others, the pilot study is conducted first, and results from that study are used to determine the sample size for the main experiment. 
At this point, data collection, analysis, and reporting proceed much as in any other experiment.

@@

# Protocol Development {#sec-exp-dev}

It would be difficult to develop a full data generating model without some idea of the experimental protocol: the basic equipment required for the experiment, some idea of what questions users will answer, where and how data will be collecteed, and so on. 
These experimental design factors are fairly natural for scientists to accumulate over the course of imagining and planning an experiment. 
When conducting graphical tests, however, there are additional considerations beyond those taught in a standard experimental design course.
Experimenters must carefully consider how much information participants should have about the experiment, what platform to use to recruit participants, and the experimental infrastructure underlying data collection. 
In addition to our brief overview of considerations in our research, @kochariConductingWebBasedExperiments2019 has many helpful suggestions for conducting web-based cognitive and perceptual studies that also apply to statistical graphics experiments.

@@

## Infrastructure

@@

When using Shiny to collect data, we store participant response data and experiment parameters using SQLite tables, which are automatically synchronized with cloud storage and tracked with version control. 
This ensures that we have incremental records of the tables during data collection, and that all data is stored across multiple locations, guarding against hardware failures. 
In some studies, data is unique to each participant; in these cases, we highly recommend saving all generated data to a database as well, so that it is possible to go back and examine responses alongside the data used to generate the graphical stimuli.
Hard drive space is extremely cheap relative to almost any other cost in an experiment; saving all of the data is a sensible measure.

## Participant Recruitment

@@

## Participant Instructions

@@

## Attention Checks and Answer Validation

@@

- Attention checks - questions inserted into the experiment which appear similarly to actual trials but which instruct participants to select a specific response or complete a trivial task (e.g. select the result of 2+2). Participants who do not successfully complete a certain percentage of attention check tasks may be excluded from the experiment. More guidance on attention checks can be found in @muszynskiAttentionChecksHow2023.

- User input validation - mechanisms which reduce the potential parameter space of user inputs to a valid or reasonable set of parameters. Examples include not allowing participants to select negative variance values, requiring continuous user-drawn lines on an interactive chart, and even checking to ensure that the user's input is of the correct type. Validation mechanisms may also ensure that participants answer all questions before proceeding to the next page.

- Time-based checks - @kochariConductingWebBasedExperiments2019 suggests that in tasks with longer instructions, it may be useful to remove participants who did not spend a certain amount of time on the instructions page. Similarly, it may be justifiable to remove participants who completed the experiment in an unreasonably short or long amount of time. On the other end of the spectrum, participants who take several hours to complete an experiment with a median completion time of 15 minutes may have experienced technical issues or been distracted.

Pre-specifying conditions where participant data will be removed from the study before analysis is critical for predictable issues, in order to streamline the analysis and defend against p-hacking concerns. 
When participants provide anomalous responses that are identified after the experiment, it is much harder, but not impossible, to justify removing the data. 
Justifications might include that no reasonable viewer would have produced a certain response, as in @sineillusion, or that the response most likely occurred due to a data entry error or a fundamental misunderstanding of the question, but this is only viable if the data is an egregious outlier.
In less severe cases, it is often preferable to leave the data in, increasing estimate variability but avoiding the need to justify and defend data cleaning decisions.

## Demographic Data

@@

## Ethics Review
Once the protocol is developed, researchers generally have enough information to get approval from the ethics board to conduct an experiment on human participants. 
We are most familiar with regulations in the United States and can speak to that general process. 
Most graphics experiments conducted in the US fall under the "exempt" category of experiments which require only basic review and approval. 
These experiments record demographic information at a level that is not individually identifiable, ask participants to complete tasks that involve no risk, and experimenters record information which would not be embarrassing to participants if exposed. 
In order to ensure that our experiments fall into this category, we will often ensure that demographic information and participant responses are collected and stored separately from any identifiable information, such as e.g. user IDs from the participant recruitment platform that allow us to monitor task completion and pay participants for their time.
Researchers should also ensure that they comply with privacy laws such as the European [General Data Protection Regulation (GDPR)](https://gdpr-info.eu/chapter-3/); complying with this law while collecting data which is not identifiable can be complicated. 
We recommend consulting with your ethics board and institutional recommendations in order to maintain legal compliance and safeguard participant privacy.

# Experimental Design {#sec-exp-design}

In statistical design terms, most of our studies involve some type of balanced incomplete block design, where participants are assigned to a subset of experimental conditions which allow for estimation of the full range of effects specified in the model. 
The particular structure of these designs depends heavily on the factorial structure of the study and the contrasts of interest. 
In some experiments, it is important that participants see each set of data only once, while in others, it is critical that participants see the same set of data represented graphically under multiple conditions, at which point it becomes necessary to manipulate the order of trials to ensure that participants do not have back-to-back trials using the same data, because that might influence the responses. 
When randomization and blocking are used in the experiment, it is often beneficial to simulate the experiment before it is conducted to ensure that the data collection platform is correctly balancing and randomizing trials. 
Of particular concern is that in online experiments with demographic controls, some categories fill faster than others. 
In extreme circumstances, this can result in imbalances in block allocation within subgroups, which may complicate analysis of demographic data. 
It is preferable to iron out these issues before the main study begins, rather than trying to patch the data collection software in the middle of the study.

It is difficult to offer specific advice on the number of participants to include in the study, because the experimental design, participant engagement level, and the specific factors under investigation have such a large effect on the number of trials one participant can reasonably complete before fatigue effects increase response variance. 
The cognitive fatigue constraints on the experimental design are important, but otherwise, power calculations for graphics experiments are similar to comparable experiments in other disciplines, in that a pilot study provides the necessary inputs to the sample size calculation for the main experiment.
Some experts have begun to recommend that instead of pilot-study informed power calculations, experimenters use a more general approach based on the intended analysis method [@brysbaertHowManyParticipants2019]. 
As power calculations for various statistical methods can be easily found in any experimental design textbook (such as @easterlingFundamentalsStatisticalExperimental2015), here we provide some basic guidelines based on past experiments.

In psychophysical graphics experiments, such as @luModelingJustNoticeable2022, there are often fewer participants with many more trials per participant (28 participants, 250 trials each); the stimuli in these experiments are often much simpler (e.g. one or two plots instead of 20 in a lineup) and engagement is limited to detection rather than estimation or prediction.
Psychophysics experiments, particularly those designed for analysis using Rasch models, require that each participant assesses a full factorial set of stimuli, whereas analysis with generalized linear mixed models allows for use of incomplete blocks and other strategies that spread the cognitive burden across several participants. 
These experiments have also been traditionally completed in person, which may also explain the large number of trials each participant is asked to complete.

Online graphics experiments often have between 300 and 600 participants [@robinsonPerceptionCognitiveImplications,], but this varies widely with the experimental design, type of stimuli used, and level of participant engagement required; @loyVariationsQQPlots2016 and @clusters had more than 1300 participants,  while @sineillusion and @vanderplasFramedReproducingRevisiting2019 had fewer than 150 participants. 
In-person experiments [e.g. @vanderplasSpatialReasoningData2016] often use fewer participants than online equivalents, but these experiments may require more tasks of each participant, reflecting the increased logistical costs of scheduling participants, supervising the experiment, and entering the data, if tasks are completed on paper. 
In some cases, in-person experiments may combat cognitive fatigue by scheduling the experiment across multiple visits; this leaves experiments vulnerable to participant drop-out effects, but can effectively balance cognitive load and participant recruitment costs in some situations.

@@

Once the data generating model is set, the protocol is developed, the experiment is designed, and ethics paperwork has been submitted, the next step is another round of testing.
The goal of the initial sequence of testing is to ensure that the experiment is set up properly and that no issues have been overlooked. 
Pilot testing also provides an opportunity to ensure that directions are clear, participants know what they are supposed to be doing, and that the designed study has sufficient power to detect an effect. 
Our studies usually go through 2-3 rounds of preliminary testing, with at least one of those rounds involving any relatives and friends who are less technologically savvy. 
We also purposely include talented individuals who can accidentally crash any testing applications, as a way to harden our data collection software before deployment.

<!-- Pilot study samples do not need to be representative of any particular population, though we would caution against using exclusively visualization researchers or statisticians in your pilot sample because some issues are less likely to show up with knowledgeable participants.  -->

One highly useful (but not strictly essential) component of an experiment that can be set up during the pilot testing stage is a basic analysis script which summarizes all data collected to date visually. 
We have used such scripts in the past to produce automatically updating dashboards or web pages, allowing for real time or near-real time monitoring of data collection efforts. 
This provides an easy way to summarize completion of the experiment so that individuals can receive credit (if using services like MTurk or Prolific), but also allows interested participants to see individual results, which can be a factor when recruiting on social media. 
As data collection online can also happen extremely quickly (300 participants in <2h in our most recent Prolific experiment), this can improve data monitoring, allowing any issues to be spotted and resolved quickly. 
If server load is a potential issue, it may also help to release batches of trials over a longer period of time in order to minimize the chance of having to make participants wait for others to complete the task before the server can handle additional connections^[This is the one major drawback to our preferred solution of self-hosting a Shiny server to handle data collection: the free version of Shiny server is limited to about 15 connections at any given time. Prolific has recently added rate-limiting functions to the experimental control  platform, which makes controlling the number of active jobs much easier.]. 
Batch trial releases can also be used to ensure that participants are recruited across different time zones; in some studies this is beneficial, while in others it may be more important to control the release time to target individuals in e.g. North America rather than Europe.

@@

## Qualitative Responses
In many cases, it is helpful to combine participants' qualitative reasoning with their quantitative responses to designed graphical experiments. 
This approach provides useful context as to what participants use to make their decisions, and can be useful when assessing why unexpected responses occurred. 
Qualitative questions can be as simple as asking for a free response explanation of the quantitative response(s), but it can also be effective to use qualitative questions to prompt participants about problem solving strategies, interesting features in the data, and more.

We have used word clouds to show overall themes in participant explanations of why specific lineup panels were selected [@clusters]; when paired with an appropriate linear model it became clear that participants were fixating on unequal cluster size as a visual cue. 
In cases where participants are provided with additional utilities such as calculators and scratchpads, it can be useful to select responses from individual participants which illustrate the different types of calculations performed, but analyzing this data quantitatively can be difficult, as it may be incomplete or difficult to code systematically.
